---
title: "Demo analysis for TraDIS infection screens with cmdstanr"
author: "Laura Jenniches"
date: "24/5/2024"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=F}
knitr::opts_chunk$set(warning=F, message=F)
```

```{r global-options, include=FALSE}
# set result directory for plots etc.
res_dir <- "../results/demo/"
data_dir <- "../data/"
if(!dir.exists("../results")) dir.create("../results")
if(!dir.exists(res_dir)) dir.create(res_dir)

# load meta data for Shigella
meta_data <- read.table("../annotations/Shigella_meta_combined.csv", header=T, sep="\t", quote="\"", stringsAsFactors=F)
seq_list <- unique(meta_data$seq)
n_seq <- length(seq_list)
meta_data$length <- meta_data$end - meta_data$start + 1
```

```{r setup, results="hide"}
## CRAN
library(ggplot2)
library(viridis)

library(tidyverse)
library(data.table)

library(cmdstanr)
library(bayesplot)

source("../R/prepare_data.R")
source("../R/analyze_stanfit.R")
source("../R/utils.R")

# Changing the font to Arial might not always work. Choose other font instead.
theme_set(theme_bw(base_size = 10))

cb_palette <- c("#864C95", "#5876A3", "#9BD0BB", "#FFD966", "#E76570", "#F0845D")
cb_palette2 <- c("#5876A3", "#F0845D", "gray", "darkred", "darkblue", "#9BD0BB")
cb_palette3 <- c("#424242", "#d36c6c")
```


Load demo count data. counts_pseudo_demo.csv contains 50 selected (neutral) pseudo genes. counts_all_demo.csv contains 50 selected genes for which the fitness scores will be determined.

```{r}
lab <- "pseudo_demo"
counts_file <- paste0(data_dir, "counts_", lab, ".csv.gz")
data_table_pseudo_demo <- fread(counts_file, data.table=F)

lab <- "all_demo"
counts_file <- paste0(data_dir, "counts_", lab, ".csv.gz")
data_table_demo <- fread(counts_file, data.table=F)

sample_labs <- colnames(data_table_demo)[grep("INPUT|OUTPUT", colnames(data_table_demo))]
sample_stats <- sample_stats_shigella(sample_labs, sample_labs %>% factor(levels=sample_labs) %>% as.numeric)
```

Create dfit files which contain the information required for the stan model.

```{r}
lab <- "all_demo"
create_dfit(data_table_demo, sample_stats, lab, description="", res_dir)

lab <- "pseudo_demo"
create_dfit(data_table_pseudo_demo, sample_stats, lab, description="", res_dir)
```

Create data list for cmdstanr (normalization).

```{r}
lab_in <- "INPUT"
lab_out <- "OUTPUT"
lab <- "pseudo_demo"
dfit_file <- paste0(res_dir, "dfit_", lab, ".csv")

# import data prepared for cmdstanr
dfit_NZ <- fread(dfit_file, data.table=F)

# for neutral (pseudo) genes: calculate number of non-zero (N_nz) insertion sites in output, 
sample_stats <- subset(dfit_NZ, type=="pseudo") %>% group_by(exp_id, sample) %>%
  summarise(N_nz=length(which(raw_out>0)), n_g=length(raw),
            N_in=mean(N), N_out=mean(N_out))
# fraction of pseudo genes with zero count in output as compared to input
sample_stats$Z <- (sample_stats$n_g-sample_stats$N_nz)/sample_stats$n_g

# tables with input and output raw counts with correctly ordered genes and samples
raw_counts_in <- t(data_table_pseudo_demo[, sample_stats$sample])
raw_counts_out <- t(data_table_pseudo_demo[, sub(lab_in, lab_out, sample_stats$sample)])
  
# gene id
it_g <- data_table_pseudo_demo$it_g
# total number of transposon insertions
N_ti <- length(it_g)
# total number of genes (should be 50 in demo)
N_g <- max(it_g)
# fraction of zero counts in pseudo genes
Z_ps <- sample_stats$Z
# number of input-output pairs
N_exp <- length(Z_ps)
# calculate sample statistics
sample_stats <- dfit_NZ %>% group_by(exp_id, sample) %>% summarise(N_nz=length(which(raw_out>0)), n_g=length(raw), N_in=mean(N), N_out=mean(N_out))
# maximum number of insertion sites per gene
N_ti_max <- max(sample_stats$n_g)

data_list <- list(
  N_ti=N_ti,
  N_g=N_g,
  N_exp=N_exp,
  Z_ps=Z_ps,
  it_g=it_g,
  raw_counts_in=raw_counts_in,
  raw_counts_out=raw_counts_out,
  N_ti_max=N_ti_max
)
```

Run cmdstan and save fit results. Alternatively, data can be exported with stan_rdump (see example_cmdstan.Rmd).
Make sure cmdstan is installed: https://mc-stan.org/cmdstanr/articles/cmdstanr.html

```{r}
# check_cmdstan_toolchain()
# install_cmdstan(cores = 2)
stan_file <- paste0(getwd(), "/../stan_models/ZINB_pair2_norm.stan")
mod <- cmdstan_model(stan_file, pedantic=F)

json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)

# fit model with cmdstan
fit0 <- mod$sample(data = json_file, chains = 2, adapt_delta = 0.95, parallel_chains = 2, iter_warmup = 1000, iter_sampling = 1000, max_treedepth = 12)

# save fitted model
fit0$save_object(file = paste0(res_dir, "fit_norm.RDS"))
rm(fit0)
```

Bayesian diagnostics

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_norm.RDS"))

# use the bayesplot packages to check for divergent transitions and correlations between parameters
# Pairs plots
x <- fit0$draws()
i <- "norm_1"
p <- mcmc_pairs(x, pars = paste0("inv_rho_g[", 1:5, "]"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()
i <- "norm_2"
p <- mcmc_pairs(x, pars = c("mu_rho", "mu", "sigma_rho"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()

# trace plots to check whether the model fit has converged and whether all chains agree
p <- mcmc_trace(x, pars = paste0("inv_rho_g[", 1:6, "]"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "trace_", i, ".pdf", sep=""), width = 7, height = 4)
print(p)
dev.off()
png(paste(res_dir, "trace_", i, ".png", sep=""), width = 400, height = 250)
print(p)
dev.off()

i <- "norm"
p <- mcmc_nuts_energy(nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "nuts_energy_", i, ".pdf", sep=""), width = 7, height = 3)
print(p)
dev.off()
png(paste(res_dir, "nuts_energy_", i, ".png", sep=""), width = 400, height = 200)
print(p)
dev.off()
```

Extract technical parameters from ZINB_pair2_norm.stan

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_norm.RDS"))

sample_labs <- colnames(data_table_demo)[grep("INPUT|OUTPUT", colnames(data_table_demo))]
sample_stats <- sample_stats_shigella(sample_labs, sample_labs %>% factor(levels=sample_labs) %>% as.numeric)

sample_stats <- sample_stats %>% arrange(exp_id)
norm_fac <- fit0$draws(variables=c("norm_fac"), format="matrix") %>% apply(2, median)
sample_stats$norm_fac <- rep(norm_fac, each=2)
a_in <- fit0$draws(variables=c("a"), format="matrix") %>% apply(2, median)
b <- fit0$draws(variables=c("b"), format="matrix") %>% apply(2, median)
```

Export full dataset (Shigella_all_data.R) to res_dir and save normalized read counts in a long dataframe (dfit_NZ_norm.csv). Run the Stan model ZINB_pair2_logFC.stan with cmdstan and copy the resulting csv file to data_dir. Alternatively, use our cmdstan results provided in the data_dir directory.

```{r}
lab_in <- "INPUT"
lab_out <- "OUTPUT"
lab <- "all_demo"
dfit_file <- paste0(res_dir, "dfit_", lab, ".csv")

# import data prepared for cmdstanr
dfit_NZ <- fread(dfit_file, data.table=F)
dfit_NZ$n_ps <- sample_stats$norm_fac[match(dfit_NZ$sample, sample_stats$sample)]
dfit_NZ_pseudo <- fread(paste0(res_dir, "dfit_pseudo_demo.csv"))

# for neutral (pseudo) genes: calculate number of non-zero (N_nz) insertion sites in output, 
sample_stats <- dfit_NZ_pseudo %>% group_by(exp_id, sample) %>%
  summarise(N_nz=length(which(raw_out>0)), n_g=length(raw),
            N_in=mean(N), N_out=mean(N_out))
# fraction of pseudo genes with zero count in output as compared to input
sample_stats$Z <- (sample_stats$n_g-sample_stats$N_nz)/sample_stats$n_g

# tables with input and output raw counts with correctly ordered genes and samples
raw_counts_in <- t(data_table_demo[, sample_stats$sample])
raw_counts_out <- t(data_table_demo[, sub(lab_in, lab_out, sample_stats$sample)])
  
# gene id
it_g <- data_table_demo$it_g
# total number of transposon insertions
N_ti <- length(it_g)
# total number of genes (should be 50 in demo)
N_g <- max(it_g)
# fraction of zero counts in pseudo genes
Z_ps <- sample_stats$Z
# number of input-output pairs
N_exp <- length(Z_ps)
# calculate sample statistics
sample_stats <- dfit_NZ %>% group_by(exp_id, sample) %>% summarise(N_nz=length(which(raw_out>0)), n_g=length(raw), N_in=mean(N), N_out=mean(N_out))
# maximum number of insertion sites per gene
N_ti_max <- max(sample_stats$n_g)

data_list <- list(
  N_ti=N_ti,
  N_g=N_g,
  N_exp=N_exp,
  Z_ps=Z_ps,
  it_g=it_g,
  raw_counts_in=raw_counts_in,
  raw_counts_out=raw_counts_out,
  N_ti_max=N_ti_max,
  norm_fac=norm_fac,
  a_in=a_in,
  b=b
)
```

```{r}
# check_cmdstan_toolchain()
# install_cmdstan(cores = 2)
stan_file <- paste0(getwd(), "/../stan_models/ZINB_pair2_logFC.stan")
mod <- cmdstan_model(stan_file, pedantic=F)

json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)

# fit model with cmdstan
fit0 <- mod$sample(data = json_file, chains = 2, adapt_delta = 0.95, parallel_chains = 2, iter_warmup = 1000, iter_sampling = 1000, max_treedepth = 12)

# save fitted model
fit0$save_object(file = paste0(res_dir, "fit_logFC.RDS"))
rm(fit0)
```

Bayesian diagnostics

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_logFC.RDS"))

# use the bayesplot packages to check for divergent transitions and correlations between parameters
# Pairs plots
x <- fit0$draws()
i <- "logFC_1"
p <- mcmc_pairs(x, pars = c("mu_rho", paste0("inv_rho_g[", 1:4, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()
i <- "logFC_2"
p <- mcmc_pairs(x, pars = c("mu", paste0("logFC_g[", 1:4, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), , width = 400, height = 400)
print(p)
dev.off()

# trace plots to check whether the model fit has converged and whether all chains agree
p <- mcmc_trace(x, pars = c("mu", paste0("logFC_g[", 1:5, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "trace_", i, ".pdf", sep=""), width = 7, height = 4)
print(p)
dev.off()
png(paste(res_dir, "trace_", i, ".png", sep=""), width = 400, height = 250)
print(p)
dev.off()

i <- "logFC"
p <- mcmc_nuts_energy(nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "nuts_energy_", i, ".pdf", sep=""), width = 7, height = 3)
print(p)
dev.off()
png(paste(res_dir, "nuts_energy_", i, ".png", sep=""), width = 400, height = 200)
print(p)
dev.off()
```

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_logFC.RDS"))

gene_stats <- dfit_NZ %>% group_by(locus_tag, it_g) %>% summarise(n_gz=length(which(raw_out==0)), n_g=length(raw), logcpm_in=mean(log(raw*n_ps)))
gene_stats$drop_out <- gene_stats$n_gz/gene_stats$n_g

results_fit <- data.frame(
  locus_tag = data_table_demo$locus_tag %>% unique()
)
results_fit$start <- meta_data$start[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$seq <- meta_data$seq[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$length <- meta_data$length[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$type <- meta_data$type[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$com_id <- meta_data$com_id[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$description <- meta_data$description[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$n_TIS <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$n_nz <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)] - gene_stats$n_gz[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$logCPM <- gene_stats$logcpm_in[match(results_fit$locus_tag, gene_stats$locus_tag)]

results_fit$inv_rho_g <- fit0$draws(variables=c("inv_rho_g"), format="matrix") %>% apply(2, median)

results_fit$logFC <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, median)
results_fit$sd_logFC <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, sd)
logFC_pseudo <- results_fit %>% subset(type=="pseudo") %>% pull(logFC) %>% median()
results_fit$p_value <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, function(x) p_value(x, int0=0))
results_fit$bayes_factor <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, function(x) Bayes_factor(x, mu=0, sigma=5))
results_fit$z <- results_fit$logFC/results_fit$sd_logFC

results_fit <- results_fit %>% subset(n_TIS>2)
results_fit$FDR <- 1
results_fit$FDR[which(results_fit$z<=0)] <- calc_FDR_z(results_fit$z[which(results_fit$z<=0)], p0=0.7, sd=0.8)
results_fit$FDR[which(results_fit$z>0)] <- calc_FDR_z(-results_fit$z[which(results_fit$z>0)], p0=0.7, sd=0.8)
```

Some figures

```{r}
results_fit_cleaned <- results_fit
plot_df <- rbind(data.frame(z=rnorm(1e5, 0, 0.8), type="prior"),
                 results_fit_cleaned[,c("z", "type")]) %>% arrange(type)
plot_df$type <- plot_df$type %>% factor(levels=c("CDS", "pseudo", "prior", "h0"))

p_h0 <- ggplot() +
  geom_histogram(aes(x=z, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", breaks=(-30:30)/3, data=subset(plot_df, type %in% c("CDS", "pseudo"))) +
  geom_density(aes(x=z, y=0.7*nrow(results_fit_cleaned)*(..density..)/3, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", data=subset(plot_df, type=="prior")) +
  scale_color_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_fill_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_alpha_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=c(0.4, 0, 0)) +
  scale_x_continuous(name=expression(z[logFC]), limits=c(-10, 10)) +
  scale_y_continuous(name="count") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_h0)
cairo_pdf(paste(res_dir, "z_values_h0.pdf", sep=""), width = 3, height = 2.2)
print(p_h0)
dev.off()
```
```{r}
# MA plot
p_MA <- ggplot(results_fit_cleaned %>% arrange(type), aes(x=log2(exp(logCPM+logFC/2)), y=log2(exp(logFC)))) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_point(aes(col=type, alpha=type), size=0.5) +
  geom_point(aes(alpha="FDR\u22640.2", col="FDR\u22640.2"), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=0.2)) +
  geom_point(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max)), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=p_max)) +
  geom_text(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max), label=com_id), data=subset(results_fit_cleaned, FDR<=0.01), size=2.5) +
  scale_color_manual(values=cb_palette2[c(3,6,4,2)], breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_alpha_manual(values=c(0.35, 0.6, 1, 1), breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_x_continuous(name="logCPM") +
  scale_y_continuous(name="logFC") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_MA)
cairo_pdf(paste(res_dir, "logFC_abundance.pdf", sep=""), width = 3, height = 3)
print(p_MA)
dev.off()
```
