---
title: "Compute fitness scores and create figures for Shigella infection screen with the cmdstan results"
author: "Laura Jenniches"
date: "24/5/2024"
output:
  pdf_document: default
  html_document: default
---

!!!! Note !!!!
Running this script requires normalizing the data and running cmdstan first (see normalize_TraDIS_screen.Rmd). The cmdstan results are also provided in the directory "data".

```{r global-options, include=FALSE}
# set result directory for plots etc.
res_dir <- "../results/TraDIS_screen/"
norm_dir <- "../results/normalize_TraDIS_screen/"
data_dir <- "../data/"
if(!dir.exists("../results")) dir.create("../results")
if(!dir.exists(res_dir)) dir.create(res_dir)

# load meta data for Shigella
meta_data <- read.table("../annotations/Shigella_meta_combined.csv", header=T, sep="\t", quote="\"", stringsAsFactors=F)
seq_list <- unique(meta_data$seq)
n_seq <- length(seq_list)
meta_data$length <- meta_data$end - meta_data$start + 1
```

```{r setup, results="hide"}
## CRAN
library(knitr)
library(Cairo)

library(ggplot2)
library(ggrastr)
library(ggrepel)

library(tidyverse)
library(data.table)

library(bayesplot)
library(rstan)

library(extrafont)
# font_import("y")
# fonts()
loadfonts()

source("../R/prepare_data.R")
source("../R/analyze_stanfit.R")
source("../R/utils.R")

# Changing the font to Arial might not always work. Choose other font instead.
theme_set(theme_bw(base_size = 10, base_family="Arial"))

cb_palette <- c("#864C95", "#5876A3", "#9BD0BB", "#FFD966", "#E76570", "#F0845D")
cb_palette2 <- c("#5876A3", "#F0845D", "gray", "darkred", "darkblue", "#9BD0BB")
cb_palette3 <- c("#424242", "#d36c6c")
```

Load cmdstan results from ZINB_pair2_logFC.stan

Load "long" data frame with count data (one insertion site per row including input/output counts and meta data like normalization factor).

```{r}
lab <- "all_Bnorm"
dfit_NZ <- fread(paste0(norm_dir, "dfit_NZ_norm.csv"), quote=F, data.table=F)
results_samples <- rbind(
  cbind(fread(cmd=sprintf('zcat %s | grep -v "#"', paste(data_dir, "Shigella_", lab, "_1.csv.gz", sep="")), quote=F, data.table=F), Chain=1),
  cbind(fread(cmd=sprintf('zcat %s | grep -v "#"', paste(data_dir, "Shigella_", lab, "_2.csv.gz", sep="")), quote=F, data.table=F), Chain=2)
) # for data from external cmdstan run
```

Calculate sampling median of technical parameters and add this information to long count data frame.

```{r}
gamma_sa <- apply(results_samples[, grep("gamma", colnames(results_samples))], 2, median)
a <- apply(results_samples[, c("a.1", "a.2")], 2, median)
dfit_NZ$gamma_sa <- gamma_sa[match(dfit_NZ$exp_id, 1:42)]
dfit_NZ$theta_g <- (dfit_NZ$gamma_sa + a[1] + a[2]*(dfit_NZ$alpha_in-dfit_NZ$alpha_sa)) %>% sapply(logit_m1)
```

Assign mean log-counts, fitness scores (log-fold changes), and gene-wise dispersion to locus tags. Retrieve order of genes saved during normalization (counts_all.csv).

Calculate Bayesian p-values, Bayes' factors and z-scores.

```{r}
locus_tags <- fread(paste0(norm_dir, "counts_all.csv"), data.table=F) %>% pull("locus_tag") %>% unique
gene_stats <- dfit_NZ %>% group_by(locus_tag, it_g) %>% summarise(n_gz=length(which(raw_out==0)), n_g=length(raw), logcpm_in=mean(log(raw*n_ps)))
gene_stats$drop_out <- gene_stats$n_gz/gene_stats$n_g

results_fit <- data.frame(
  locus_tag = locus_tags
)
results_fit$start <- meta_data$start[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$seq <- meta_data$seq[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$length <- meta_data$length[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$type <- meta_data$type[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$com_id <- meta_data$com_id[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$description <- meta_data$description[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$n_TIS <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$n_nz <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)] - gene_stats$n_gz[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$logCPM <- gene_stats$logcpm_in[match(results_fit$locus_tag, gene_stats$locus_tag)]

results_fit$inv_rho_g <- apply(results_samples[, grep("inv_rho_g", colnames(results_samples))], 2, median)

results_fit$logFC <- apply(results_samples[, grep("logFC_g", colnames(results_samples))], 2, median)
results_fit$sd_logFC <- apply(results_samples[, grep("logFC_g", colnames(results_samples))], 2, sd)
logFC_pseudo <- results_fit %>% subset(type=="pseudo") %>% pull(logFC) %>% median()
results_fit$p_value <- apply(results_samples[, grep("logFC_g", colnames(results_samples))] %>% as.matrix(), 2, function(x) p_value(x, int0=0))
results_fit$bayes_factor <- apply(results_samples[, grep("logFC_g", colnames(results_samples))] %>% as.matrix(), 2, function(x) Bayes_factor(x, mu=0, sigma=5))
results_fit$z <- results_fit$logFC/results_fit$sd_logFC
```

Plot the distribution of the number of insertion sites per gene as a pie plot.

```{r}
results_fit_nTI <- results_fit
results_fit_nTI$n_TIS[which(results_fit_nTI$n_TIS>=20)] <- 20
results_fit_nTI <- results_fit_nTI %>% group_by(n_TIS) %>% summarise(value=length(locus_tag))
results_fit_nTI <- data.frame(n_TIS=0, value=meta_data %>% subset(!(locus_tag %in% results_fit$locus_tag)) %>% nrow) %>% rbind(results_fit_nTI)
results_fit_nTI$value_sum <- sapply(1:nrow(results_fit_nTI), function(i) sum(results_fit_nTI$value[1:i]) - 0.5*results_fit_nTI$value[i])
results_fit_nTI$lab <- results_fit_nTI$n_TIS
results_fit_nTI$lab[which(results_fit_nTI$n_TIS==20)] <- 
  paste0("≥20",
         ":\n",
         results_fit_nTI$value[which(results_fit_nTI$n_TIS==20)],
         " genes")
results_fit_nTI$x <- 1.55
results_fit_nTI$x[which(results_fit_nTI$n_TIS==20)] <- 1

p_pi <- results_fit_nTI %>% ggplot() +
  geom_bar(aes(x=1, y=value, fill=n_TIS), col="black", stat="identity", alpha=1) +
  geom_text(aes(x=x, y=value_sum, label=lab), size=2.5) +
  geom_text(aes(x=1, y=value_sum, label=paste0((value/nrow(meta_data)*100) %>% round(), " %")), size=2.5, data=subset(results_fit_nTI, n_TIS==0)) +
  coord_polar("y") +
  scale_fill_gradientn(name="#TIS", colours=cb_palette[1:4]) +
  theme_minimal(base_size=10) +
  theme(
    rect = element_rect(fill = "transparent"),
    legend.position = 'none',
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title=element_text(face="bold"),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-0.5,4), "cm")
  )
print(p_pi)
cairo_pdf(paste(res_dir, "n_TIS_pi.pdf", sep=""), width = 2.2, height = 2.2)
print(p_pi)
dev.off()
```

Add FDR to z-scores by comparing to a standard normal distribution. See function calc_FDR_z in analyze_stanfit.R for details.

```{r}
results_fit <- results_fit %>% subset(n_TIS>2)
results_fit$FDR <- 1
results_fit$FDR[which(results_fit$z<=0)] <- calc_FDR_z(results_fit$z[which(results_fit$z<=0)], 0.9)
results_fit$FDR[which(results_fit$z>0)] <- calc_FDR_z(-results_fit$z[which(results_fit$z>0)], 0.9)

theta_sa <- exp(gamma_sa)/(1+exp(gamma_sa))
gamma_med <- gamma_sa %>% median()
theta_med <- exp(gamma_med)/(1+exp(gamma_med))
```

Abundance dependent drop-out

We accounted for stochastic loss in our ZINB model. Here, we compare the calculated loss per input abundance bin to the model result. The curve is not expected to fit to the data points exactly since the observed loss is a combination of gene essentiality and bottleneck effects.

```{r}
breaks=(-20:20)/2
dfit_NZ_binned <- dfit_NZ %>% mutate(bin_mu=cut(alpha_in-alpha_sa, breaks=breaks))
sum_mu <- dfit_NZ_binned %>% group_by(bin_mu) %>% summarise(height_mu=length(bin_mu))
dfit_dropout_binned <- dfit_NZ %>% subset(raw_out==0) %>% mutate(bin_mu=cut(alpha_in-alpha_sa, breaks=breaks))
sum_mu_out <- dfit_dropout_binned %>% group_by(bin_mu) %>% summarise(height=length(bin_mu))
sum_mu$height_dropout_mu <- sum_mu_out$height[match(sum_mu$bin_mu, sum_mu_out$bin_mu)]
sum_mu$height_dropout_mu[is.na(sum_mu$height_dropout_mu)] <- 0
bin_map <- data.frame(bin=cut(breaks, breaks=breaks), values=breaks)
sum_mu$bin <- bin_map$values[match(sum_mu$bin_mu, bin_map$bin)]
line_df <- data.frame(x=(-150:150)/30)
line_df$p <- gamma_med + a[1]+ a[2]*line_df$x
line_df$y <- exp(line_df$p)/(1+exp(line_df$p))

p_drop <- ggplot(sum_mu) +
  geom_point(aes(x=bin, y=height_dropout_mu/height_mu, col="data", alpha="data"), size=1) +
  geom_line(aes(x=x, y=y, col="fit", alpha="fit"), data=line_df) +
  scale_color_manual(name="", values=c(cb_palette[1], "black")) +
  scale_alpha_manual(name="", values=c(0.6, 1)) +
  scale_x_continuous(name=expression(paste(alpha,"-",bar(alpha))), limits=c(-5,2)) +
  scale_y_continuous(name="dropout", breaks=c(0,0.5,1)) + #, sec.axis = sec_axis( trans=~.*3000, name="abundance")) +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_drop)
cairo_pdf(paste(res_dir, "dropout_abundance.pdf", sep=""), width = 3, height = 2.2)
print(p_drop)
dev.off()
```

Distribution of z-scores compared to standard normal distribution.

```{r}
results_fit_cleaned <- results_fit
plot_df <- rbind(data.frame(z=rnorm(1e5, 0, 0.9), type="prior"),
                 results_fit_cleaned[,c("z", "type")]) %>% arrange(type)
plot_df$type <- plot_df$type %>% factor(levels=c("CDS", "pseudo", "prior", "h0"))
p_h0_pseudo <- ggplot() +
  geom_histogram(aes(x=z, fill=type, col=type, alpha=type), linetype="solid", breaks=(-30:30)/3, data=subset(plot_df, type %in% c("CDS", "pseudo"))) +
  geom_density(aes(x=z, y=0.87*nrow(results_fit_cleaned)*(..density..)/3, fill=type, col=type, alpha=type), linetype="solid", data=subset(plot_df, type=="prior")) +
  scale_color_manual(name="", breaks=c("CDS", "pseudo", "prior"), labels=c("CDS", "pseudo", "h0"), values=c("transparent", "transparent", cb_palette[5])) +
  scale_fill_manual(name="", breaks=c("CDS", "pseudo", "prior"), labels=c("CDS", "pseudo", "h0"), values=cb_palette[c(1,4,5)]) +
  scale_alpha_manual(name="", breaks=c("CDS", "pseudo", "prior"), labels=c("CDS", "pseudo", "h0"), values=c(0.5, 0.5, 0)) +
  scale_x_continuous(name=expression(z[logFC]), limits=c(-10, 10)) +
  scale_y_continuous(name="count") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_h0_pseudo)
cairo_pdf(paste(res_dir, "z_values_h0_pseudo.pdf", sep=""), width = 3, height = 2.2)
print(p_h0_pseudo)
dev.off()

p_h0 <- ggplot() +
  geom_histogram(aes(x=z, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", breaks=(-30:30)/3, data=subset(plot_df, type %in% c("CDS", "pseudo"))) +
  geom_density(aes(x=z, y=0.85*nrow(results_fit_cleaned)*(..density..)/3, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", data=subset(plot_df, type=="prior")) +
  scale_color_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_fill_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_alpha_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=c(0.4, 0, 0)) +
  scale_x_continuous(name=expression(z[logFC]), limits=c(-10, 10)) +
  scale_y_continuous(name="count") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_h0)
cairo_pdf(paste(res_dir, "z_values_h0.pdf", sep=""), width = 3, height = 2.2)
print(p_h0)
dev.off()
```

Export result tables with log2CPM values, log2-fold changes, z-scores and FDR. Plot FDR vs. z-score.

```{r}
# convert log-fold changes to log2-fold changes
results_export <- results_fit_cleaned %>% mutate(log2FC=log2(exp(logFC)), log2CPM=log2(exp(logCPM)))
# calculate median log2CPM
fwrite(results_export %>% mutate(log2CPM=log2CPM+log2FC/2) %>% dplyr::select(c("locus_tag", "com_id", "description", "type", "length", "n_TIS", "n_nz", "log2CPM", "log2FC", "z", "FDR")), paste0(res_dir, "results_ZINB_pair_logFC.csv"))

fwrite(results_export %>% mutate(log2CPM=log2CPM+log2FC/2)  %>% subset(locus_tag %in% subset(meta_data, seq=="CP037923.1")$locus_tag) %>% dplyr::select(c("locus_tag", "com_id", "description", "type", "length", "n_TIS", "n_nz", "log2CPM", "log2FC", "z", "FDR")), paste0(res_dir, "results_ZINB_pair_logFC_chrom.csv"))
fwrite(results_export %>% mutate(log2CPM=log2CPM+log2FC/2)  %>% subset(locus_tag %in% subset(meta_data, seq=="CP037924.1")$locus_tag) %>% dplyr::select(c("locus_tag", "com_id", "description", "type", "length", "n_TIS", "n_nz", "log2CPM", "log2FC", "z", "FDR")), paste0(res_dir, "results_ZINB_pair_logFC_plasmid.csv"))


p_FDR <- ggplot() +
  geom_point(aes(x=z, y=-log10(FDR), col="CDS"), size=0.75, alpha=0.5, data=results_fit_cleaned %>% subset(FDR>1e-10 & type=="CDS")) +
  geom_point(aes(x=z, y=-log10(FDR), col="pseudo"), size=0.5, alpha=0.5, data=results_fit_cleaned %>% subset(FDR>1e-10 & type=="pseudo")) +
  scale_color_manual(name="", values=c(cb_palette[c(1,4)])) +
  geom_hline(yintercept=c(1), linetype="dashed", size=0.75) +
  xlab(expression(z[logFC])) +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_FDR)
cairo_pdf(paste(res_dir, "z_FDR.pdf", sep=""), width = 3, height = 2.2)
print(p_FDR)
dev.off()
```

Example: insertion sites in gene EKN05_000820 across all input and output libraries.

```{r}
dfit_NZ <- dfit_NZ %>% mutate(alpha_in=log(raw*n_ps+0.01), alpha_out=log(raw_out+0.01))
dfit_NZ$com_id <- meta_data$com_id[match(dfit_NZ$locus_tag, meta_data$locus_tag)]
dfit_stats <- dfit_NZ %>% group_by(locus_tag, com_id) %>% summarise(alpha_g=mean(alpha_out), zeros=sum(theta_g), med_out=mean(raw_out))
gene <- "EKN05_000820"
x_start <- subset(meta_data, com_id==gene)$start
x_end <- subset(meta_data, com_id==gene)$end
dfit_NZ_gene <- subset(dfit_NZ, com_id==gene)
alpha_gene <- subset(dfit_stats, com_id==gene)$alpha_g
exp_zeros <- sum(dfit_NZ_gene$theta_g)
act_zeros <- subset(results_fit, com_id==gene) %>% mutate(n_z=n_TIS-n_nz) %>% pull(n_z)

p_raw <- dfit_NZ_gene %>% ggplot(aes(x=pos)) +
  geom_bar(aes(y=raw, fill="input", col="input", alpha="input"), stat="identity", position="dodge", width=20) +
  geom_bar(aes(y=raw_out, fill="output", col="output", alpha="output"), stat="identity", position="dodge", width=20) +
  facet_grid(exp_id~.) +
  scale_fill_manual(name="", values=cb_palette3[c(1,2)]) +
  scale_color_manual(name="", values=cb_palette3[c(1,2)]) +
  scale_x_continuous(name="", limits=c(x_start, x_end), breaks=c()) +
  scale_y_continuous(name="counts", breaks=c(0, 5000)) +
  scale_alpha_manual(name="", values=c(1, 1)) +
  ggtitle("Raw TIS") +
  theme(rect=element_rect(fill="transparent"),
        plot.title=element_text(size=11),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), strip.text.y.right = element_text(angle = 0),
        axis.text=element_text(color = "black"), panel.grid=element_blank())
print(p_raw)
cairo_pdf(paste(res_dir, "raw_TIS.pdf", sep=""), width = 1.5, height = 2.4)
print(p_raw)
dev.off()
```

Stochastic (technical) loss for all insertion sites in gene EKN05_000820. Input and output counts for gene EKN05_000820 after normalization and correction for stochastic loss.

```{r}
x <- (-150:150)/30
gamma_min <- dfit_NZ_gene$gamma_sa %>% min
gamma_max <- dfit_NZ_gene$gamma_sa %>% max
line_df <- data.frame(x=rep(x, 3), p=c(gamma_min + a[1] + a[2]*x, gamma_med + a[1]+ a[2]*x, gamma_max + a[1] + a[2]*x), type=rep(c("min", "med", "max"), each=length(x)))
line_df$y <- exp(line_df$p)/(1+exp(line_df$p))

p_drop_gene <- ggplot(dfit_NZ_gene) +
  geom_line(aes(x=x, y=y, col="fit", linetype=(type!="med"), group=type), data=line_df) +
  geom_point(aes(x=alpha_in-alpha_sa, y=theta_g, col="data", alpha=1-theta_g)) +
  scale_color_manual(name="", values=cb_palette3[1:2], guide=F) +
  scale_alpha_continuous(name="", range=c(0.3, 1)) +
  scale_linetype_manual(values=c("solid", "dashed"), guide=F) +
  scale_x_continuous(name=expression(paste(alpha["rgk,in"],"-",bar(alpha[r]))), limits=c(-2,2)) +
  scale_y_continuous(name=expression(theta[rgk]), breaks=c(0,0.5,1)) + #, sec.axis = sec_axis( trans=~.*3000, name="abundance")) +
  ggtitle("Technical dropout") +
  theme(rect=element_rect(fill="transparent"),
        plot.title=element_text(size=11),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="none", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_drop_gene)

p_norm <- dfit_NZ_gene %>% ggplot(aes(x=pos)) +
  geom_bar(aes(y=raw*n_ps, fill="input", alpha=1-theta_g), col="transparent", stat="identity", position="dodge", width=20) +
  geom_bar(aes(y=raw_out, fill="output", alpha=1), col="transparent", stat="identity", position="dodge", width=20) +
  scale_fill_manual(name="", values=cb_palette3[c(1,2)]) +
  scale_x_continuous(name="", limits=c(x_start, x_end), breaks=c()) +
  scale_y_continuous(name="counts") +
  scale_alpha_continuous(guide="none", range=c(0.3,1)) +
  ggtitle("Normalized counts") +
  theme(rect=element_rect(fill="transparent"),
        plot.title=element_text(size=11),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_norm)

it_g <- which(locus_tags==subset(meta_data, com_id==gene)$locus_tag)
p_pos <- data.frame(logFC=results_samples[, paste("logFC_g", it_g, sep=".")]) %>% ggplot() +
  geom_density(aes(logFC), fill=cb_palette3[2], alpha=0.5, col="transparent") +
  geom_vline(xintercept=0, linetype="dashed") +
  xlab("logFC") +
  ylab("") +
  ggtitle("Posterior distribution") +
  theme(rect=element_rect(fill="transparent"),
        plot.title=element_text(size=11),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
```

```{r}
load(paste0(norm_dir, "p_corr_zoom.Rdata"))
p_model <- cowplot::plot_grid(
  p_corr_zoom,
  NULL,
  p_raw,
  NULL,
  cowplot::plot_grid(NULL, p_drop_gene, p_norm, ncol=1, rel_heights=c(0.05, 0.5, 0.45)),
  NULL,
  cowplot::plot_grid(NULL, p_pos, NULL, ncol=1, rel_heights=c(0.15, 0.5, 0.15)),
  ncol=7, rel_widths=c(0.8,0.15,1,0.15,0.9,0.15,0.8), scale=1)
print(p_model)
cairo_pdf(paste(res_dir, "model_", gene, ".pdf", sep=""), width = 6.9, height = 2.5)
print(p_model)
dev.off()
```

MA plot
Gene-wise log2-fold change vs. mean log2CPM (mean of input and output).

```{r}
# MA plot
p_MA <- ggplot(results_fit_cleaned %>% arrange(type), aes(x=log2(exp(logCPM+logFC/2)), y=log2(exp(logFC)))) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_point(aes(col=type, alpha=type), size=0.5) +
  geom_point(aes(alpha="FDR\u22640.2", col="FDR\u22640.2"), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=0.2)) +
  geom_point(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max)), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=p_max)) +
  geom_text_repel(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max), label=com_id), data=subset(results_fit_cleaned, FDR<=0.01), size=2.5) +
  scale_color_manual(values=cb_palette2[c(3,6,4,2)], breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_alpha_manual(values=c(0.35, 0.6, 1, 1), breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_x_continuous(name="logCPM") +
  scale_y_continuous(name="logFC") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_MA)
cairo_pdf(paste(res_dir, "logFC_abundance.pdf", sep=""), width = 3, height = 3)
print(p_MA)
dev.off()
```

MA plot for chromosomal genes.

```{r}
p_MA_chrom <- ggplot(results_fit_cleaned %>% subset(seq=="CP037923.1") %>% arrange(type), aes(x=log2(exp(logCPM+logFC/2)), y=log2(exp(logFC)))) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_point(aes(col=type, alpha=type), size=1) +
  geom_point(aes(alpha="adv. FDR\u22640.2", col="adv. FDR\u22640.2"), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=0.2 & seq=="CP037923.1" & logFC<0)) +
  geom_point(aes(alpha="detr. FDR\u22640.2", col="detr. FDR\u22640.2"), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=0.2 & seq=="CP037923.1" & logFC>0)) +
  geom_point(aes(alpha=paste0("adv. FDR\u2264", p_max), col=paste0("adv. FDR\u2264", p_max)), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=p_max & seq=="CP037923.1" & logFC<0)) +
  geom_point(aes(alpha=paste0("detr. FDR\u2264", p_max), col=paste0("detr. FDR\u2264", p_max)), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=p_max & seq=="CP037923.1" & logFC>0)) +
  geom_text_repel(aes(label=com_id), min.segment.length=unit(0, 'lines'), nudge_x=-0.3, nudge_y=-1, data=subset(results_fit_cleaned, grepl("acrA|acrB|mnmE|mnmG|tolC|trmH", com_id) & seq=="CP037923.1"), size=2.5) +
  scale_color_manual(name="", values=cb_palette2[c(3,6,4,5,2,1)], breaks=c("CDS", "pseudo", "adv. FDR\u22640.2", "detr. FDR\u22640.2", paste0("adv. FDR\u2264", p_max), paste0("detr. FDR\u2264", p_max), paste0("FDR\u2264", p_max))) +
  scale_alpha_manual(name="", values=c(0.35, 0.6, 1, 1, 1, 1), breaks=c("CDS", "pseudo", "adv. FDR\u22640.2", "detr. FDR\u22640.2", paste0("adv. FDR\u2264", p_max), paste0("detr. FDR\u2264", p_max), paste0("FDR\u2264", p_max))) +
  scale_x_continuous(name="logCPM") +
  scale_y_continuous(name="logFC") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_MA_chrom)
cairo_pdf(paste(res_dir, "logFC_abundance_chrom.pdf", sep=""), width = 3, height = 3)
print(p_MA_chrom)
dev.off()
```

MA plot for virulence plasmid.

```{r}
p_MA_plas <- ggplot(results_fit_cleaned %>% subset(seq=="CP037924.1") %>% arrange(type), aes(x=log2(exp(logCPM+logFC/2)), y=log2(exp(logFC)))) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_point(aes(col=type, alpha=type), size=1) +
  geom_point(aes(alpha="adv. FDR\u22640.2", col="adv. FDR\u22640.2"), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=0.2 & seq=="CP037924.1" & logFC<0)) +
  geom_point(aes(alpha="detr. FDR\u22640.2", col="detr. FDR\u22640.2"), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=0.2 & seq=="CP037924.1" & logFC>0)) +
  geom_point(aes(alpha=paste0("adv. FDR\u2264", p_max), col=paste0("adv. FDR\u2264", p_max)), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=p_max & seq=="CP037924.1" & logFC<0)) +
  geom_point(aes(alpha=paste0("detr. FDR\u2264", p_max), col=paste0("detr. FDR\u2264", p_max)), shape=1, size=1.5, data=subset(results_fit_cleaned, FDR<=p_max & seq=="CP037924.1" & logFC>0)) +
  geom_text_repel(aes(label=com_id), data=subset(results_fit_cleaned, FDR<=0.01 & seq=="CP037924.1" & logFC<0), size=2.5) +
  scale_color_manual(name="", values=cb_palette2[c(3,6,4,5,2,1)], breaks=c("CDS", "pseudo", "adv. FDR\u22640.2", "detr. FDR\u22640.2", paste0("adv. FDR\u2264", p_max), paste0("detr. FDR\u2264", p_max), paste0("FDR\u2264", p_max))) +
  scale_alpha_manual(name="", values=c(0.35, 0.6, 1, 1, 1, 1), breaks=c("CDS", "pseudo", "adv. FDR\u22640.2", "detr. FDR\u22640.2", paste0("adv. FDR\u2264", p_max), paste0("detr. FDR\u2264", p_max), paste0("FDR\u2264", p_max))) +
  scale_x_continuous(name="logCPM") +
  scale_y_continuous(name="logFC") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_MA_plas)
cairo_pdf(paste(res_dir, "logFC_abundance_plasmid.pdf", sep=""), width = 3, height = 3)
print(p_MA_plas)
dev.off()
```

Chromosomal positions of (significant) log2FC

```{r}
dfit_NZ <- dfit_NZ %>% mutate(bin=ceiling(pos/500)*500)
dfit_NZ$bin[which(dfit_NZ$seq=="CP037924")] <- dfit_NZ$bin[which(dfit_NZ$seq=="CP037924")]
dfit_int <- dfit_NZ %>% group_by(seq, bin) %>% summarise(height_in=sum(alpha_in))
base_data <- dfit_int %>% group_by(seq) %>% summarise(x_min=min(bin), x_max=max(bin))
base_data$x_min <- 0

offset <- 0.1
results_fit_cleaned$x_coord <- results_fit_cleaned$start
results_fit_cleaned$logFC_plot <- log2(exp(results_fit_cleaned$logFC))
results_fit_cleaned$logFC_plot[which(results_fit_cleaned$logFC<0)] <- results_fit_cleaned$logFC_plot[which(results_fit_cleaned$logFC<0)] - offset
results_fit_cleaned$logFC_plot[which(results_fit_cleaned$logFC>0)] <- results_fit_cleaned$logFC_plot[which(results_fit_cleaned$logFC>0)] + offset
results_fit_cleaned$ymin <- offset
results_fit_cleaned$ymin[which(results_fit_cleaned$logFC<0)] <- -offset

p_pos_chrom <- ggplot(results_fit_cleaned %>% subset(seq=="CP037923.1")) +
  geom_segment(aes(x=x_coord, y=ymin, xend=x_coord, yend=logFC_plot, col=(logFC>0))) +
  geom_segment(aes(x=x_coord, y=ymin, xend=x_coord, yend=logFC_plot), col="red", data=subset(results_fit_cleaned, seq=="CP037923.1" & FDR <= p_max), inherit.aes=F) +
  coord_polar() +
  scale_color_manual(values=c(cb_palette[1], "gray", "gray")) +
  theme_minimal(base_size=10) +
  scale_x_continuous(breaks=(0:4)*1e6) +
  ylim(c(-10, 1)) +
  theme(
    rect=element_rect(fill="transparent"),
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    plot.title=element_text(face="bold"),
    panel.grid.minor = element_blank(),
    plot.margin = unit(rep(-1,4), "cm")
  ) +
  geom_segment(data=base_data %>% subset(seq=="CP037923"), aes(x = x_min, y = 0, xend = x_max, yend = 0, col=seq), alpha=0.8, size=0.3, inherit.aes = FALSE)
print(p_pos_chrom)
cairo_pdf(paste(res_dir, "genome_logFC.pdf", sep=""), width = 4.5, height = 4.5)
print(p_pos_chrom)
dev.off()
```

```{r}
offset <- 0.1
p_pos_plas <- ggplot(results_fit_cleaned %>% subset(seq=="CP037924.1")) +
  geom_segment(aes(x=x_coord, y=ymin, xend=x_coord, yend=logFC_plot, col=(logFC>0))) +
  geom_segment(aes(x=x_coord, y=ymin, xend=x_coord, yend=logFC_plot), col="red", data=subset(results_fit_cleaned, seq=="CP037924.1" & FDR <= p_max), inherit.aes=F) +
  coord_polar() +
  scale_color_manual(values=c(cb_palette[1], "gray", "gray")) +
  theme_minimal() +
  scale_x_continuous(breaks=(0:2)*1e5) +
  ylim(c(-10, 1)) +
  theme(
    rect=element_rect(fill="transparent"),
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    plot.title=element_text(face="bold"),
    panel.grid.minor = element_blank(),
    plot.margin = unit(rep(-1,4), "cm")
  ) +
  geom_segment(data=base_data %>% subset(seq=="CP037924"), aes(x = 0, y = 0, xend = x_max, yend = 0, col=seq), alpha=0.8, size=0.3, inherit.aes = FALSE)
print(p_pos_plas)
cairo_pdf(paste(res_dir, "genome_logFC_plasmid.pdf", sep=""), width = 4.5, height = 4.5)
print(p_pos_plas)
dev.off()
```

Diagnostics for the MCMC sampling, pair plots can inform about correlations between parameters and divergent transitions.

```{r}
np <- extract_np(results_samples, n_iter=1000)
```

```{r}
plot_pairs(paste(res_dir, "pairs_hyperparameters.pdf", sep=""), results_samples, np, c("a.1", "mu_rho", "sigma_rho", "inv_rho_g.1"))
```

```{r}
plot_pairs(paste(res_dir, "pairs_logFC.pdf", sep=""), results_samples, np, c("inv_rho_g.1", "logFC_g.1", "inv_rho_g.2294", "logFC_g.2294"))
```
