---
title: "Shigella pilot experiments"
author: "Laura Jenniches"
date: "17/4/2024"
output:
  pdf_document: default
  html_document: default
---

```{r global-options, include=FALSE}
# set result directory for plots etc.
res_dir <- "../results/pilot_experiments/"
data_dir <- "../data/"
if(!dir.exists("../results")) dir.create("../results")
if(!dir.exists(res_dir)) dir.create(res_dir)

## Bioconductor
library('edgeR')

## CRAN
library(ggplot2)
library(dplyr)
library(viridis)
library(data.table)
library(extrafont)
# font_import("y")
# fonts()
loadfonts()

# Changing the font to Arial might not always work. Choose other font instead.
theme_set(theme_bw(base_size = 10, base_family="Arial"))

source("../R/utils.R")
source("../R/normalization.R")

cb_palette <- c("#864C95", "#5876A3", "#9BD0BB", "#FFD966", "#E76570")
cb_palette2 <- c("#5876A3", "#F0845D", "gray", "darkred", "darkblue", "#9BD0BB")
cb_palette3 <- c("#424242", "#d36c6c")
```

```{r}
# load gene meta data
meta_data <- fread("../annotations/Shigella_meta_combined.csv", quote="\"", stringsAsFactors=F, data.table=F)
row.names(meta_data) <- meta_data$locus_tag
meta_data$it_g <- as.numeric(factor(meta_data$locus_tag))
meta_data$length <- meta_data$end - meta_data$start + 1
seq_list <- unique(meta_data$seq)
n_seq <- length(seq_list)
```

load count data for L10/L30 sub-libraries with 10,000 and 30,000 transposon insertion sites (TIS), respectively

```{r}
data_table <- fread(paste(data_dir, "counts_TraDIS_L10_L30.csv.gz", sep=""), quote="\"", stringsAsFactors=F, data.table=F)
read_counts <- data_table[, grep("rpt", colnames(data_table))]

cond_data <- sapply(colnames(read_counts), function(x) strsplit(as.character(x), "[-]")[[1]])
cond <- data.frame(lib.type=sub("L", "", cond_data[2,]),
                   sample=cond_data[1,],
                   condition=cond_data[3,],
                   replicate=as.integer(sub("rpt", "", cond_data[4,])),
                   strain="M90T",
                   identifier=colnames(read_counts),
                   row.names=colnames(read_counts),
                   stringsAsFactors=F)

N_sa <- nrow(cond)
```

```{r}
data_table$it_g <- meta_data$it_g[match(data_table$locus_tag, meta_data$locus_tag)]
rownames(read_counts) <- paste(data_table$seq, data_table$pos, sep="_")
pseudo_genes <- subset(read_counts, data_table$type=="pseudo")

# drop insert sites which have 0 counts accross all libraries
keep <- which(rowSums(read_counts) > 0)
read_counts <- read_counts[keep,]
data_table <- data_table[keep,]

rownames(read_counts) <- data_table$TI
pseudo_genes <- subset(read_counts, data_table$type=="pseudo")
```

normalization with pseudo genes

```{r}
lib_sizes <- colSums(read_counts)
norm_facs <- get_norm_facs_nonzero(read_counts, pseudo_genes, count_min=1, sample_min=3)
write.table(data.frame(nf=norm_facs), paste(res_dir, "norm_facs_L10_L30.csv", sep=""), row.names=T, quote=F)
```

Estimate number of samples required for TraDIS infection screen

```{r}
insertion_sites_in <- fread(paste(data_dir, "counts_pilot_TraDIS.csv.gz", sep=""), data.table = F, stringsAsFactors = F) %>% dplyr::select(matches("Min|locus"))
dfit_min <- data.frame(locus_tag=insertion_sites_in$locus_tag, counts_sum=rowSums(insertion_sites_in[1:3])) %>% subset(counts_sum>1)
estimate_N_samples <- data.frame(N_samples=(5:50)*2)
estimate_N_samples$n_genes_min3 <- sapply(estimate_N_samples$N_samples, function(x) {
  data.frame(sel=sample(dfit_min$locus_tag, x*3000, replace=T, prob=dfit_min$counts_sum)) %>% subset(sel!="inter") %>%
    group_by(sel) %>% summarise(n_TIS=n()) %>% subset(n_TIS>2) %>% nrow()
})
estimate_N_samples$n_genes_min1 <- sapply(estimate_N_samples$N_samples, function(x) {
  data.frame(sel=sample(dfit_min$locus_tag, x*1000, replace=T, prob=dfit_min$counts_sum)) %>% subset(sel!="inter") %>%
    group_by(sel) %>% summarise(n_TIS=n()) %>% subset(n_TIS>2) %>% nrow()
})
estimate_N_samples$n_genes_min2 <- sapply(estimate_N_samples$N_samples, function(x) {
  data.frame(sel=sample(dfit_min$locus_tag, x*2000, replace=T, prob=dfit_min$counts_sum)) %>% subset(sel!="inter") %>%
    group_by(sel) %>% summarise(n_TIS=n()) %>% subset(n_TIS>2) %>% nrow()
})

# plot number of genes with at least 3 insertion sites for 1,000/2,000/3,000 insertion sites per sub-library
p_N_samples <- estimate_N_samples %>% ggplot(aes(N_samples)) +
  geom_point(aes(y=n_genes_min1, col="1 k")) +
  geom_point(aes(y=n_genes_min2, col="2 k")) +
  geom_point(aes(y=n_genes_min3, col="3 k")) +
  scale_x_continuous(name="no. samples", limits=c(0,100)) +
  scale_y_continuous(name="no. genes with min. 3 TIS") +
  scale_color_manual(name="mutants per lib.", values=cb_palette) +
  geom_vline(xintercept=40, linetype="dashed") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.25,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_rect(fill=NA, size=0.5), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_N_samples)
cairo_pdf(paste(res_dir, "N_samples_min3_TIS.pdf", sep=""), width = 3, height = 2.2, bg="transparent")
print(p_N_samples)
dev.off()
```

Calculate the number of insertion sites per gene for 40 parallel infections with 3,000 insertion sites each (in the input)

```{r, eval=F}
df_40samples_list <- lapply(1:100, function(i) {
  run_df <- data.frame(sel=sample(dfit_min$locus_tag, 40*3000, replace=T, prob=dfit_min$counts_sum)) %>% subset(sel!="inter") %>%
    group_by(sel) %>% summarise(n_TIS=n())
  run_df$n_TIS[which(run_df$n_TIS>=20)] <- 20
  n_noTIS <- meta_data %>% subset(!(locus_tag %in% run_df$sel)) %>% nrow
  rbind(run_df %>% group_by(n_TIS) %>% summarise(count=n()), data.frame(n_TIS=0, count=n_noTIS))})

df_40samples <- do.call(rbind, df_40samples_list) %>% group_by(n_TIS) %>% summarise(value=mean(count))
df_40samples$value_sum <- sapply(1:nrow(df_40samples), function(i) sum(df_40samples$value[1:i]) - 0.5*df_40samples$value[i])
df_40samples$lab <- df_40samples$n_TIS
df_40samples$lab[which(df_40samples$n_TIS==20)] <- 
  paste0("≥20",
         ":\n",
         round(df_40samples$value[which(df_40samples$n_TIS==20)]),
         " genes")
df_40samples$x <- 1.55
df_40samples$x[which(df_40samples$n_TIS %in% c(20))] <- 1
fwrite(df_40samples, paste0(data_dir, "sim40samples_run2.csv"), quote=T)
```

Plot the number of insertion sites per gene for 40 parallel infections with 3,000 insertion sites each (in the input)

```{r}
df_40samples <- fread(paste0(data_dir, "sim40samples.csv"), data.table=F, quote="\"")
p_pi <- df_40samples %>% ggplot() +
  geom_bar(aes(x=1, y=value, fill=n_TIS), col="black", stat="identity", alpha=1) +
  geom_text(aes(x=x, y=value_sum, label=lab), size=2.25) +
  geom_text(aes(x=1, y=df_40samples$value_sum[which(df_40samples$n_TIS==0)],
                label=paste0(round(100*df_40samples$value[which(df_40samples$n_TIS==0)]/nrow(meta_data))," %")), size=2.25) +
  coord_polar("y") +
  scale_fill_gradientn(name="#TIS", colours=cb_palette[1:4]) +
  theme_minimal() +
  theme(
    rect=element_rect(fill="transparent"),
    legend.position = 'none',
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title=element_text(face="bold"),
    panel.grid = element_blank(),
    plot.margin = unit(rep(-0.5,4), "cm")
  )
print(p_pi)
cairo_pdf(paste(res_dir, "n_TIS_40_samples.pdf", sep=""), width = 2.2, height = 2.2, bg="transparent")
print(p_pi)
dev.off()
```

create long data frame from L10/L30 TraDIS libraries

```{r}
dfit <- data.frame(gene=character(), it_g=integer(), type=character(), k=integer(), seq=character(), pos=integer(), sample=character(), raw=numeric(), stringsAsFactors=F)
dfit <- rbind(dfit, as.data.frame(reshape::melt(cbind(read_counts, data_table[(N_sa+1):(ncol(data_table))]), id=c("locus_tag", "it_g", "type", "k", "seq", "pos"))))
colnames(dfit) <- c("gene", "it_g","type","k", "seq", "pos", "sample", "raw")
dfit$lib.type <- cond$lib.type[match(dfit$sample, cond$identifier)]
dfit$condition <- cond$condition[match(dfit$sample, cond$identifier)]
dfit$replicate <- cond$replicate[match(dfit$sample, cond$identifier)]
dfit$TI <- paste(dfit$seq, dfit$pos, sep="_")
dfit$com_id <- meta_data$com_id[match(dfit$gene, meta_data$locus_tag)]
dfit$nf <- norm_facs[match(dfit$sample, names(norm_facs))]
dfit$N <- lib_sizes[match(dfit$sample, names(lib_sizes))]
dfit$cpm <- (dfit$raw+0.5)/(dfit$nf*dfit$N+1)*1e+06
dfit$logcpm <- log((dfit$raw+0.5)/(dfit$nf*dfit$N+1)*1e+06)
dfit$it_sa <- as.numeric(dfit$sample)
dfit_pseudo <- subset(dfit, type=="pseudo")
```

Combine counts of corresponding minimally enriched input (Min) and output (Out) insertions sites

```{r}
dfit_comp <- subset(dfit, grepl("Min", sample)) %>% subset(!duplicated(paste0(TI,lib.type)))
dfit_comp <- rbind(cbind(dfit_comp, rep=1), cbind(dfit_comp, rep=2), cbind(dfit_comp, rep=3)) %>% arrange(lib.type)
dfit_comp$replicate <- dfit_comp$rep
dfit_comp$sample <- paste(dfit_comp$lib.type, dfit_comp$condition, dfit_comp$replicate, sep="_")
dfit_out <- subset(dfit, condition=="Out")
dfit_out$sample <- paste(dfit_out$lib.type, dfit_out$condition, dfit_out$replicate, sep="_")
dfit_comp$raw_out <- dfit_out$raw[match(paste(dfit_comp$TI, dfit_comp$lib.type, dfit_comp$replicate, sep="_"), paste(dfit_out$TI, dfit_out$lib.type, dfit_out$replicate, sep="_"))]
dfit_comp <- subset(dfit_comp, raw>9)
dfit_stats <-  dfit_comp %>% group_by(sample, replicate) %>% summarize(nTI_in=length(raw), nTI_out=length(which(raw_out>1)))
dfit_stats_pseudo <- dfit_comp %>% subset(type=="pseudo") %>% group_by(sample, replicate) %>% summarize(nTI_in=length(raw), nTI_out=length(which(raw_out>1)))
```

Figure 1 D
Plot number of insertion sites with more than 10 read counts for each library

```{r}
# keep only insertion sites with more than 10 read counts
dfit_NZ_sum <- dfit %>% subset(raw>=10) %>% group_by(lib.type, replicate, condition) %>% summarise(uTI=length(TI))
dfit_NZ_sum$lib.type <- paste("Lib.", dfit_NZ_sum$lib.type, "k")
dfit_NZ_sum$id <- paste0(dfit_NZ_sum$condition, "_R", dfit_NZ_sum$replicate)
dfit_NZ_sum$id <- factor(dfit_NZ_sum$id, levels=dfit_NZ_sum$id %>% unique())
x_lab <- dfit_NZ_sum$replicate[match(dfit_NZ_sum$id %>% unique, dfit_NZ_sum$id)]
dfit_NZ_sum$condition <- factor(dfit_NZ_sum$condition, levels=c("Min", "Inp", "Out"))

# all libraries
p_size_min <- ggplot(dfit_NZ_sum) +
  geom_col(aes(x=replicate, y=uTI, fill=condition), stat="identity", position=position_dodge(width=1, preserve = "single"), alpha=1, size=0) +
  facet_wrap(~lib.type) +
  scale_fill_manual(name="", values=cb_palette[c(2,1,3)]) +
  scale_y_continuous(name="# unique IS", labels=fancy_scientific) +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_rect(fill=NA, size=0.5), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_size_min)
cairo_pdf(paste(res_dir, "pilot_exp_unique_IS_SI.pdf", sep=""), width = 3, height = 2.2, bg="transparent")
print(p_size_min)
dev.off()

# remove minimal input for main figure
p_size <- ggplot(dfit_NZ_sum %>% subset(condition!="Min")) +
  geom_col(aes(x=replicate, y=uTI, fill=condition), stat="identity", position=position_dodge(width=1, preserve = "single"), alpha=1, size=0) +
  facet_wrap(~lib.type) +
  scale_fill_manual(name="", values=cb_palette[c(1,3)]) +
  scale_y_continuous(name="nr. unique IS", labels=fancy_scientific) +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_rect(fill=NA, size=0.5), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_size)
cairo_pdf(paste(res_dir, "pilot_exp_unique_IS.pdf", sep=""), width = 3, height = 2.2, bg="transparent")
print(p_size)
dev.off()

cairo_pdf(paste(res_dir, "pilot_S1.pdf", sep=""), width = 6.9, height = 2.2, bg="transparent")
cowplot::plot_grid(p_size_min, p_N_samples, p_pi, rel_widths = c(0.35, 0.35, 0.25), scale=c(0.95, 0.95, 0.73), ncol=3)
dev.off()
```

Figure 1E
Vary bottleneck size between 100 and 6000, sample corresponding number of insertion sites from minimally enriched input samples, compare to actual number of TIS in output samples

```{r}
counts_in <- read_counts[, grep("L10-Min", colnames(read_counts))]
counts_in <- t(t(counts_in)/colSums(counts_in))*1e6
rownames(counts_in) <- rownames(read_counts)
counts_in <- counts_in[rowSums(counts_in)>0, ]
median_counts_in <- rowMedians(counts_in)

bottleneck <- (1:60)*100
bottleneck_df <-
  data.frame(bottleneck,
             nTI_out_10=sapply(bottleneck,
                               function(size) sample(rownames(counts_in), size, replace=T, prob=median_counts_in) %>% unique %>% length)
)

counts_in <- read_counts[, grep("L30-Min", colnames(read_counts))]
counts_in <- t(t(counts_in)/colSums(counts_in))*1e6
rownames(counts_in) <- rownames(read_counts)
counts_in <- counts_in[rowSums(counts_in)>0, ]
median_counts_in <- rowMedians(counts_in)

bottleneck_df <- 
  cbind(bottleneck_df,
        nTI_out_30=sapply(bottleneck,
                          function(size) sample(rownames(counts_in), size, replace=T, prob=median_counts_in) %>% unique %>% length)
)

p_bottle <- ggplot(bottleneck_df) +
  geom_tile(aes(x, y, width=1500), fill="gray", data=data.frame(x=c(2000), y=c(0, Inf))) +
  geom_line(aes(x=bottleneck, y=nTI_out_10, col="10k")) +
  geom_line(aes(x=bottleneck, y=nTI_out_30, col="30k")) +
  geom_hline(aes(yintercept=nTI_out, col="10k", linetype=replicate %>% as.character()), data=dfit_stats %>% subset(grepl("10", sample))) +
  geom_hline(aes(yintercept=nTI_out, col="30k", linetype=replicate %>% as.character()), data=dfit_stats %>% subset(grepl("30", sample))) +
  scale_color_manual(name="", values=cb_palette[c(1,4)]) +
  scale_linetype_manual(name="Rep", values=c(2,2,2), guide=F) +
  scale_y_continuous(name="nr. unique output IS") +
  scale_x_continuous(name="bottleneck size") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_rect(fill=NA, size=0.5), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_bottle)
cairo_pdf(paste(res_dir, "bottleneck.pdf", sep=""), width = 3, height = 2.2, bg="transparent")
print(p_bottle)
dev.off()
```

Figure 1H
Compare enrichment of TraDIS libraries at 37 vs. 30C with RNA-seq of WT Shigella at 37 vs 30C

```{r, fig.show="asis"}
tt_37vs30 <- read.table("../results/pilot_RNAseq_edgeR/tt_M90T_37vs30C.csv", header=T, sep="\t", quote="")
tt_enr37vs30 <- read.table("../results/pilot_TraDIS_edgeR/tt_Enr_37_30.csv", header=T, sep="\t", quote="")
tt_enr_30 <- read.table("../results/pilot_TraDIS_edgeR/tt_Enr_Sub_30.csv", header=T, sep="\t", quote="")
tt_enr_37 <- read.table("../results/pilot_TraDIS_edgeR/tt_Enr_Sub_37.csv", header=T, sep="\t", quote="")
cor_virF <- data.frame(locus_tag=tt_37vs30$locus_tag,
                       logFC_M90T_37vs30=tt_37vs30$logFC,
                       FDR_M90T_37vs30=tt_37vs30$FDR,
                       logFC_enr37vs30=tt_enr37vs30$logFC[match(tt_37vs30$locus_tag, tt_enr37vs30$locus_tag)], 
                       FDR_enr37vs30=tt_enr37vs30$FDR[match(tt_37vs30$locus_tag, tt_enr37vs30$locus_tag)], 
                       gene_name=meta_data$com_id[match(tt_37vs30$locus_tag, meta_data$locus_tag)],
                       seq=meta_data$seq[match(tt_37vs30$locus_tag, meta_data$locus_tag)],
                       description=tt_37vs30$description, type=tt_37vs30$type)
write.table(cor_virF, paste0(res_dir, "cor_37vs30_enr37vs30.csv"), row.names=F, quote=F, sep="\t")

cor_virF <- subset(cor_virF, seq=="CP037924.1" | gene_name %in% c("virF", "cpxR", "EKN05_024540"))
cor_virF$reg <- "neutral"
cor_virF$reg[which(cor_virF$FDR_enr37vs30<=0.01 & cor_virF$logFC_enr37vs30>1)] <- "Detr"
cor_virF$reg[which(cor_virF$FDR_enr37vs30<=0.01 & cor_virF$logFC_enr37vs30< -1)] <- "Adv"

p_corr <- ggplot(cor_virF %>% subset((FDR_M90T_37vs30<=0.01 & type=="CDS") | gene_name %in% c("virF", "cpxR", "EKN05_024540")), aes(x=logFC_M90T_37vs30, y=logFC_enr37vs30)) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_vline(xintercept=0, linetype="dashed") +
  geom_point(aes(col=reg, alpha=reg)) +
  ggrepel::geom_text_repel(aes(label=ifelse(FDR_enr37vs30<=0.05, gene_name, "")), col="black", size=2,
                           subset(cor_virF, gene_name %in% c("virF", "cpxR", "EKN05_024540", "ipaA", "ipaD", "fis", "EKN05_024560", "virB"))) +
  scale_color_manual(name="", values=c(cb_palette2,"gray")) +
  scale_alpha_manual(name="", values=c(1, 1, 0.5)) +
  scale_x_continuous(name="RNA-seq M90T, 37\u00B0C vs. 30\u00B0C", limits=c(-2,8)) +
  scale_y_continuous(name="TraDIS enr. 37\u00B0C vs. enr. 30\u00B0C", limits=c(-3, 4)) +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_rect(fill=NA, size=0.5), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_corr)
cairo_pdf(paste(res_dir, "TraDIS_RNAseq.pdf", sep=""), width = 3, height = 2.2, bg="transparent")
print(p_corr)
dev.off()
```

Figure 1G - Venn Diagrams
Overlap between TraDIS and RNA-seq pilot experiments

```{r}
tt_virF_37 <- read.table("../results/pilot_RNAseq_edgeR/tt_M90TvsFd_37C.csv", header=T, sep="\t", quote="")

tt_virF_30 <- read.table("../results/pilot_RNAseq_edgeR/tt_M90TvsFd_30C.csv", header=T, sep="\t", quote="")
virF37_genes <- tt_virF_37 %>% subset(FDR<=0.01 & abs(logFC)>1 & type=="CDS") %>% pull(locus_tag)
# remove the gene which is in none of the other categories for plotting
virF30_genes <- tt_virF_30 %>% subset(FDR<=0.01 & abs(logFC)>1 & type=="CDS") %>% pull(locus_tag)
temp_genes <- tt_37vs30 %>% subset(FDR<=0.01 & abs(logFC)>1 & type=="CDS") %>% pull(locus_tag)

temp <- VennDiagram::venn.diagram(x = list(virF37_genes, temp_genes, virF30_genes),
                          category.names = c(paste0("virF regulated genes\n(M90T vs ΔvirF\n 37\u00B0C) ", length(virF37_genes)),
                                             paste0("Temperature regulated genes\n(M90T\n 37\u00B0C vs. 30\u00B0C) ", length(temp_genes)),
                                             paste0("M90T vs ΔvirF\n 30\u00B0C ", length(virF30_genes))),
                          filename = NULL,
                          fill = cb_palette[c(1,3,4)],
                          fontfamily = "sans",
                          cat.fontfamily = "sans",
                          col="transparent",
                          # imagetype = "svg",
                          margin=0.1,
                          cex=0.7,
                          cat.cex=0.7,
                          output=F)
grid::grid.draw(temp)

cairo_pdf(paste0(res_dir, 'virF_temp.pdf'), bg="transparent")
grid::grid.draw(temp)
dev.off()
```

```{r}
p_bar <- rbind(data.frame(gene=length(tt_enr_37 %>% subset(FDR<=0.01 & logFC< -1 & type=="CDS") %>% pull(locus_tag)), type="Adv", tmp="37"),
      data.frame(gene=length(tt_enr_37 %>% subset(FDR<=0.01 & logFC>1 & type=="CDS") %>% pull(locus_tag)), type="Detr", tmp="37"),
      data.frame(gene=length(tt_enr_30 %>% subset(FDR<=0.01 & logFC< -1 & type=="CDS") %>% pull(locus_tag)), type="Adv", tmp="30"),
      data.frame(gene=length(tt_enr_30 %>% subset(FDR<=0.01 & logFC>1 & type=="CDS") %>% pull(locus_tag)), type="Detr", tmp="30")) %>%
  ggplot(aes(x=tmp, y=gene, fill=type)) +
  geom_col(stat="identity", position=position_dodge(width=1, preserve = "single"), alpha=1, size=0) +
  # facet_wrap(~tmp, strip.position="bottom") +
  scale_fill_manual(name="", values=cb_palette2) +
  scale_x_discrete(name="temp [°C]") +
  scale_y_continuous(name="nr. mutants") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_bar)
cairo_pdf(paste(res_dir, "Enr_37vs30.pdf", sep=""), width = 1.8, height = 2.2, bg="transparent")
print(p_bar)
dev.off()
```

```{r}
p_pilot <- cowplot::plot_grid(
  cowplot::plot_grid(p_size, p_bottle, p_bar, NULL, scale=c(0.9, 0.9, 0.9, 0.9), rel_widths=c(0.44, 0.37, 0.21, 0.022), ncol=4, labels=c("D", "E", "F",""), label_size=14, label_fontface="plain"),
  cowplot::plot_grid(NULL, temp, NULL, p_corr, scale=c(0.95, 0.9), rel_widths=c(0.05, 0.36, 0.05, 0.54), ncol=4), ncol=1, rel_heights=c(0.45, 0.55), labels=c("", "G", "", "H"), label_size=14, label_fontface="plain")
print(p_pilot)
cairo_pdf(paste(res_dir, "pilot_1.pdf", sep=""), width = 6.9, height = 4.2, bg="transparent")
print(p_pilot)
dev.off()
```
