---
title: "Demo analysis for TraDIS infection screens with cmdstanr"
author: "Laura Jenniches"
date: "08/1/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, warning=F}
knitr::opts_chunk$set(warning=F, message=F, echo=F)
```

```{r}
# set result directory for plots etc.
res_dir <- "../results/demo/"
data_dir <- "../data/"
if(!dir.exists("../results")) dir.create("../results")
if(!dir.exists(res_dir)) dir.create(res_dir)

# load meta data for Shigella
meta_data <- read.table("../annotations/Shigella_meta_combined.csv", header=T, sep="\t", quote="\"", stringsAsFactors=F)
seq_list <- unique(meta_data$seq)
n_seq <- length(seq_list)
meta_data$length <- meta_data$end - meta_data$start + 1
```

```{r}
## CRAN
library(ggplot2)
library(viridis)

library(tidyverse)
library(data.table)

library(cmdstanr)
library(bayesplot)
source("../R/prepare_data.R")
source("../R/analyze_stanfit.R")
source("../R/utils.R")

# Changing the font to Arial might not always work. Choose other font instead.
theme_set(theme_bw(base_size = 10))

cb_palette <- c("#864C95", "#5876A3", "#9BD0BB", "#FFD966", "#E76570", "#F0845D")
cb_palette2 <- c("#5876A3", "#F0845D", "gray", "darkred", "darkblue", "#9BD0BB")
cb_palette3 <- c("#424242", "#d36c6c")
```


1) Load demo count data. counts_pseudo_demo.csv contains 200 selected (neutral) pseudo genes of ten input-output pairs for normalization. counts_all_demo.csv contains 200 selected genes (100 with a fitness effect and 100 neutral) for which the fitness scores will be determined.

```{r}
lab <- "pseudo_demo"
counts_file <- paste0(data_dir, "counts_", lab, ".csv.gz")
data_table_pseudo_demo <- fread(counts_file, data.table=F)

lab <- "all_demo"
counts_file <- paste0(data_dir, "counts_", lab, ".csv.gz")
data_table_demo <- fread(counts_file, data.table=F)

sample_labs <- colnames(data_table_demo)[grep("INPUT|OUTPUT", colnames(data_table_demo))]
sample_stats <- sample_stats_shigella(sample_labs, sample_labs %>% factor(levels=sample_labs) %>% as.numeric)
```

2) Create dfit files which contain the information required for the stan model.

```{r}
lab <- "all_demo"
create_dfit(data_table_demo, sample_stats, lab, description="", res_dir)

lab <- "pseudo_demo"
create_dfit(data_table_pseudo_demo, sample_stats, lab, description="", res_dir)
```

3) Create data list for cmdstanr (normalization).

```{r}
lab_in <- "INPUT"
lab_out <- "OUTPUT"
lab <- "pseudo_demo"
dfit_file <- paste0(res_dir, "dfit_", lab, ".csv")

# import data prepared for cmdstanr
dfit_NZ <- fread(dfit_file, data.table=F)

# for neutral (pseudo) genes: calculate number of non-zero (N_nz) insertion sites in output, 
sample_stats <- subset(dfit_NZ, type=="pseudo") %>% group_by(exp_id, sample) %>%
  summarise(N_nz=length(which(raw_out>0)), n_g=length(raw),
            N_in=mean(N), N_out=mean(N_out))
# fraction of pseudo genes with zero count in output as compared to input
sample_stats$Z <- (sample_stats$n_g-sample_stats$N_nz)/sample_stats$n_g

# tables with input and output raw counts with correctly ordered genes and samples
raw_counts_in <- t(data_table_pseudo_demo[, sample_stats$sample])
raw_counts_out <- t(data_table_pseudo_demo[, sub(lab_in, lab_out, sample_stats$sample)])
  
# gene id
it_g <- data_table_pseudo_demo$it_g
# total number of transposon insertions
N_ti <- length(it_g)
# total number of genes (should be 50 in demo)
N_g <- max(it_g)
# fraction of zero counts in pseudo genes
Z_ps <- sample_stats$Z
# number of input-output pairs
N_exp <- length(Z_ps)
# calculate sample statistics
sample_stats <- dfit_NZ %>% group_by(exp_id, sample) %>% summarise(N_nz=length(which(raw_out>0)), n_g=length(raw), N_in=mean(N), N_out=mean(N_out))
# maximum number of insertion sites per gene
N_ti_max <- max(sample_stats$n_g)

data_list <- list(
  N_ti=N_ti,
  N_g=N_g,
  N_exp=N_exp,
  Z_ps=Z_ps,
  it_g=it_g,
  raw_counts_in=raw_counts_in,
  raw_counts_out=raw_counts_out,
  N_ti_max=N_ti_max
)
```

4) Run cmdstan and save fit results. Alternatively, data can be exported with stan_rdump and the sampling can be performed with cmdstan.
Make sure cmdstan is installed: https://mc-stan.org/cmdstanr/articles/cmdstanr.html

```{r}
# This block may be skipped when running the script multiple times (eval=F).
# Run the first two commands to verify whether cmdstan is installed and to install cmdstan if necessary.
# check_cmdstan_toolchain()
# install_cmdstan(cores = 2)
stan_file <- paste0(getwd(), "/../stan_models/ZINB_pair2_norm.stan")
mod <- cmdstan_model(stan_file, pedantic=F)

json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)

# fit model with cmdstan
fit0 <- mod$sample(data = json_file, chains = 2, adapt_delta = 0.95, parallel_chains = 2, iter_warmup = 1000, iter_sampling = 1000, max_treedepth = 12)

# save fitted model
fit0$save_object(file = paste0(res_dir, "fit_norm.RDS"))
rm(fit0)
```

5) Bayesian diagnostics: pair and trace plots, NUTS energy plots.
There may be some divergent transitions (pair and trace plots) because fitting a hierarchical Bayesian model to a small number of genes is suboptimal. This results in an inefficient exploration of the parameter space (NUTS energy plot). For a full dataset, there should be no divergent transitions. This step can be accelerated when initializing the parameters adequately. 

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_norm.RDS"))

# use the bayesplot packages to check for divergent transitions and correlations between parameters
# Pairs plots
x <- fit0$draws()
i <- "norm_1"
p <- mcmc_pairs(x, pars = paste0("inv_rho_g[", 1:5, "]"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()
i <- "norm_2"
p <- mcmc_pairs(x, pars = c("mu_rho", "mu", "sigma_rho"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()

# trace plots to check whether the model fit has converged and whether all chains agree
p <- mcmc_trace(x, pars = paste0("inv_rho_g[", 1:6, "]"),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "trace_", i, ".pdf", sep=""), width = 7, height = 4)
print(p)
dev.off()
png(paste(res_dir, "trace_", i, ".png", sep=""), width = 400, height = 250)
print(p)
dev.off()

i <- "norm"
p <- mcmc_nuts_energy(nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "nuts_energy_", i, ".pdf", sep=""), width = 7, height = 3)
print(p)
dev.off()
png(paste(res_dir, "nuts_energy_", i, ".png", sep=""), width = 400, height = 200)
print(p)
dev.off()
```

6) Extract technical (normalization) parameters from fitting ZINB_pair2_norm. With the demo dataset, the following results should be optained:
a_in $\approx$ (0.05, -0.68),
b $\approx$ -0.35.

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_norm.RDS"))

sample_labs <- colnames(data_table_demo)[grep("INPUT|OUTPUT", colnames(data_table_demo))]
sample_stats <- sample_stats_shigella(sample_labs, sample_labs %>% factor(levels=sample_labs) %>% as.numeric)

sample_stats <- sample_stats %>% arrange(exp_id)
norm_fac <- fit0$draws(variables=c("norm_fac"), format="matrix") %>% apply(2, median)
sample_stats$norm_fac <- rep(norm_fac, each=2)
a_in <- fit0$draws(variables=c("a"), format="matrix") %>% apply(2, median)
b <- fit0$draws(variables=c("b"), format="matrix") %>% apply(2, median)
```

7) Create data list for cmdstanr to calculate fitness scores (logFCs) for the 200 genes demo dataset.

```{r}
lab_in <- "INPUT"
lab_out <- "OUTPUT"
lab <- "all_demo"
dfit_file <- paste0(res_dir, "dfit_", lab, ".csv")

# import data prepared for cmdstanr
dfit_NZ <- fread(dfit_file, data.table=F)
dfit_NZ$n_ps <- sample_stats$norm_fac[match(dfit_NZ$sample, sample_stats$sample)]
dfit_NZ_pseudo <- fread(paste0(res_dir, "dfit_pseudo_demo.csv"))

# for neutral (pseudo) genes: calculate number of non-zero (N_nz) insertion sites in output, 
sample_stats <- dfit_NZ_pseudo %>% group_by(exp_id, sample) %>%
  summarise(N_nz=length(which(raw_out>0)), n_g=length(raw),
            N_in=mean(N), N_out=mean(N_out))
# fraction of pseudo genes with zero count in output as compared to input
sample_stats$Z <- (sample_stats$n_g-sample_stats$N_nz)/sample_stats$n_g

# tables with input and output raw counts with correctly ordered genes and samples
raw_counts_in <- t(data_table_demo[, sample_stats$sample])
raw_counts_out <- t(data_table_demo[, sub(lab_in, lab_out, sample_stats$sample)])
  
# gene id
it_g <- data_table_demo$it_g
# total number of transposon insertions
N_ti <- length(it_g)
# total number of genes (should be 50 in demo)
N_g <- max(it_g)
# fraction of zero counts in pseudo genes
Z_ps <- sample_stats$Z
# number of input-output pairs
N_exp <- length(Z_ps)
# calculate sample statistics
sample_stats <- dfit_NZ %>% group_by(exp_id, sample) %>% summarise(N_nz=length(which(raw_out>0)), n_g=length(raw), N_in=mean(N), N_out=mean(N_out))
# maximum number of insertion sites per gene
N_ti_max <- max(sample_stats$n_g)

data_list <- list(
  N_ti=N_ti,
  N_g=N_g,
  N_exp=N_exp,
  Z_ps=Z_ps,
  it_g=it_g,
  raw_counts_in=raw_counts_in,
  raw_counts_out=raw_counts_out,
  N_ti_max=N_ti_max,
  norm_fac=norm_fac,
  a_in=a_in,
  b=b
)
```

8) Run cmdstan and save fit results. Alternatively, data can be exported with stan_rdump and the sampling can be performed with cmdstan.
Make sure cmdstan is installed: https://mc-stan.org/cmdstanr/articles/cmdstanr.html

```{r}
# This block may be skipped when running the script multiple times (eval=F).
stan_file <- paste0(getwd(), "/../stan_models/ZINB_pair2_logFC.stan")
mod <- cmdstan_model(stan_file, pedantic=F)

json_file <- tempfile(fileext = ".json")
write_stan_json(data_list, json_file)

# fit model with cmdstan
fit0 <- mod$sample(data = json_file, chains = 2, adapt_delta = 0.95, parallel_chains = 2, iter_warmup = 1000, iter_sampling = 1000, max_treedepth = 12)

# save fitted model
fit0$save_object(file = paste0(res_dir, "fit_logFC.RDS"))
rm(fit0)
```

9) Bayesian diagnostics: pair and trace plots, NUTS energy plots.
There may be some divergent transitions (pair and trace plots) because fitting the hierarchical model to a small number of genes is suboptimal. The divergences are not due to misspecifications because they are distributed all over the parameter space. The small number of genes also results in an inefficient exploration of the parameter space (NUTS energy plot). For a full dataset, there should be no divergent transitions. This step can be accelerated when initializing the parameters adequately. 

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_logFC.RDS"))

# use the bayesplot packages to check for divergent transitions and correlations between parameters
# Pairs plots
x <- fit0$draws()
i <- "logFC_1"
p <- mcmc_pairs(x, pars = c("mu_rho", paste0("inv_rho_g[", 1:4, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), width = 400, height = 400)
print(p)
dev.off()
i <- "logFC_2"
p <- mcmc_pairs(x, pars = c("mu", paste0("logFC_g[", 1:4, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "pairs_", i, ".pdf", sep=""), width = 7, height = 7)
print(p)
dev.off()
png(paste(res_dir, "pairs_", i, ".png", sep=""), , width = 400, height = 400)
print(p)
dev.off()

# trace plots to check whether the model fit has converged and whether all chains agree
p <- mcmc_trace(x, pars = c("mu", paste0("logFC_g[", 1:5, "]")),
           off_diag_args = list(size = 1, alpha = 0.5), np=nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "trace_", i, ".pdf", sep=""), width = 7, height = 4)
print(p)
dev.off()
png(paste(res_dir, "trace_", i, ".png", sep=""), width = 400, height = 250)
print(p)
dev.off()

i <- "logFC"
p <- mcmc_nuts_energy(nuts_params(fit0))
print(p)
cairo_pdf(paste(res_dir, "nuts_energy_", i, ".pdf", sep=""), width = 7, height = 3)
print(p)
dev.off()
png(paste(res_dir, "nuts_energy_", i, ".png", sep=""), width = 400, height = 200)
print(p)
dev.off()
```

10) Extract fitness scores (logFCs) from fit and calculate mean input expression (logCPM). The natural logarithm is used here.

```{r}
# load fitted model
fit0 <- readRDS(paste0(res_dir, "fit_logFC.RDS"))

gene_stats <- dfit_NZ %>% group_by(locus_tag, it_g) %>% summarise(n_gz=length(which(raw_out==0)), n_g=length(raw), logcpm_in=mean(log(raw*n_ps)))
gene_stats$drop_out <- gene_stats$n_gz/gene_stats$n_g

results_fit <- data.frame(
  locus_tag = data_table_demo$locus_tag %>% unique()
)
results_fit$start <- meta_data$start[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$seq <- meta_data$seq[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$length <- meta_data$length[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$type <- meta_data$type[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$com_id <- meta_data$com_id[match(results_fit$locus_tag, meta_data$locus_tag)]
results_fit$description <- meta_data$description[match(results_fit$locus_tag, meta_data$locus_tag)]

results_fit$n_TIS <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$n_nz <- gene_stats$n_g[match(results_fit$locus_tag, gene_stats$locus_tag)] - gene_stats$n_gz[match(results_fit$locus_tag, gene_stats$locus_tag)]
results_fit$logCPM <- gene_stats$logcpm_in[match(results_fit$locus_tag, gene_stats$locus_tag)]

results_fit$inv_rho_g <- fit0$draws(variables=c("inv_rho_g"), format="matrix") %>% apply(2, median)

results_fit$logFC <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, median)
results_fit$sd_logFC <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, sd)
logFC_pseudo <- results_fit %>% subset(type=="pseudo") %>% pull(logFC) %>% median()
results_fit$p_value <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, function(x) p_value(x, int0=0))
results_fit$bayes_factor <- fit0$draws(variables=c("logFC_g"), format="matrix") %>% apply(2, function(x) Bayes_factor(x, mu=0, sigma=5))
results_fit$z <- results_fit$logFC/results_fit$sd_logFC

results_fit <- results_fit %>% subset(n_TIS>2)
results_fit$FDR <- 1
results_fit$FDR[which(results_fit$z<=0)] <- calc_FDR_z(results_fit$z[which(results_fit$z<=0)], p0=0.7, sd=0.8)
results_fit$FDR[which(results_fit$z>0)] <- calc_FDR_z(-results_fit$z[which(results_fit$z>0)], p0=0.7, sd=0.8)
```

11) Compare the distribution of Z scores to the expected distribution under the null hypothesis (red line). Here, the fraction of log-fold changes in agreement with no fitness effect (null hypothesis) has been set to 0.7 because the demo dataset contains a very high fraction of genes with fitness effects. For a real dataset, this value should be around 0.9.

```{r}
results_fit_cleaned <- results_fit
plot_df <- rbind(data.frame(z=rnorm(1e5, 0, 0.8), type="prior"),
                 results_fit_cleaned[,c("z", "type")]) %>% arrange(type)
plot_df$type <- plot_df$type %>% factor(levels=c("CDS", "pseudo", "prior", "h0"))

p_h0 <- ggplot() +
  geom_histogram(aes(x=z, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", breaks=(-30:30)/3, data=subset(plot_df, type %in% c("CDS", "pseudo"))) +
  geom_density(aes(x=z, y=0.7*nrow(results_fit_cleaned)*(..density..)/3, fill=type!="prior", col=type!="prior", alpha=type!="prior"), linetype="solid", data=subset(plot_df, type=="prior")) +
  scale_color_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_fill_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=cb_palette[c(4,5,5)]) +
  scale_alpha_manual(name="", breaks=c(T, F), labels=c("data", "h0"), values=c(0.4, 0, 0)) +
  scale_x_continuous(name=expression(z[logFC]), limits=c(-10, 10)) +
  scale_y_continuous(name="count") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_h0)
cairo_pdf(paste(res_dir, "z_values_h0.pdf", sep=""), width = 3, height = 2.2)
print(p_h0)
dev.off()
```

12) The MA plot shows that the data is centered around zero, i.e. the normalization with the small demo dataset was successfull. The natural logarithm was converted to log2.

```{r}
# MA plot
p_max <- 0.1
p_MA <- ggplot(results_fit_cleaned %>% arrange(type), aes(x=log2(exp(logCPM+logFC/2)), y=log2(exp(logFC)))) +
  geom_hline(yintercept=0, linetype="dashed") +
  geom_point(aes(col=type, alpha=type), size=0.5) +
  geom_point(aes(alpha="FDR\u22640.2", col="FDR\u22640.2"), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=0.2)) +
  geom_point(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max)), shape=1, size=1, data=subset(results_fit_cleaned, FDR<=p_max)) +
  geom_text(aes(alpha=paste0("FDR\u2264", p_max), col=paste0("FDR\u2264", p_max), label=com_id), data=subset(results_fit_cleaned, FDR<=0.01), size=2.5) +
  scale_color_manual(values=cb_palette2[c(3,6,4,2)], breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_alpha_manual(values=c(0.35, 0.6, 1, 1), breaks=c("CDS", "pseudo", "FDR\u22640.2", paste0("FDR\u2264", p_max))) +
  scale_x_continuous(name="logCPM") +
  scale_y_continuous(name="logFC") +
  theme(rect=element_rect(fill="transparent"),
        axis.line=element_line(linewidth=unit(0.1, "pt")),
        legend.position="top", legend.margin=margin(t=-0.2, b=-0.2, unit="cm"),
        legend.key.width=unit(0.3,"cm"),legend.key.height=unit(0.3,"cm"),
        strip.background = element_blank(), panel.spacing = unit(0, "lines"),
        panel.grid=element_blank(), axis.text=element_text(color = "black"))
print(p_MA)
cairo_pdf(paste(res_dir, "logFC_abundance.pdf", sep=""), width = 3, height = 3)
print(p_MA)
dev.off()
```
